---
title: "Election 2021 (DEU) ·  🇩🇪"
author: "Holger Doering"
date: today
execute:
  cache: false
  messages: true
  warning: false
format:
  html:
    code-fold: true
    df-print: kable
    embed-resources: true
    toc: true
---

```{r corepacks}
library(conflicted) # create errors for function name conflicts

library(tidyverse)
conflicts_prefer(dplyr::filter, .quiet = TRUE)
library(knitr) # layout // tables in RMarkdown with kable()

ggplot2::theme_set(theme_void())
options(scipen = 999) # suppress scientific notation

library(estimatr) # robust standard errors
library(patchwork) # arrrange plots
library(sf) # plot maps and gespatial tools

crs_deu <- "+proj=lcc +lat_1=48 +lat_2=53 +lat_0=51 +lon_0=10" # LCC Germany parameters
```


# Source · 🗃

The Federal Returning Officer ("Der Bundeswahlleiter"): Bundestag election 2021

- [Results](https://www.bundeswahlleiter.de/en/bundestagswahlen/2021/ergebnisse.html) constituencies
  - © Der Bundeswahlleiter, Statistisches Bundesamt, Wiesbaden 2021
- [Geometric data](https://www.bundeswahlleiter.de/en/bundestagswahlen/2021/wahlkreiseinteilung/downloads.html) of constituencies
  - © Der Bundeswahlleiter, Statistisches Bundesamt, Wiesbaden 2020, Wahlkreiskarte für die Wahl zum 20. Deutschen Bundestag, Basis of the geological information © Geobasis-DE / BKG 2020

```{r}
btw_raw <- read_csv("data/btw21-results-party.csv")

party_level <- c("AfD", "CDU", "CSU", "FDP", "SPD", "GRÜNE", "DIE LINKE")

btw_dt <-
  btw_raw |>
  mutate(party = fct_relevel(party, party_level))

btw_tier1 <- btw_dt |> filter(level == "district", votes_type == "Erststimmen")
btw_tier2 <- btw_dt |> filter(level == "district", votes_type == "Zweitstimmen")

mp_raw <- read_csv("data/btw21-gewaehlte-wahlkreis.csv")

mp <-
  mp_raw |>
  mutate(party = fct_relevel(party, party_level))

wk_shp <-
  read_rds("data/btw21-districts-geometry.rds") |>
  left_join(mp |> distinct(wkr_nr, region))
wk_ctr <- read_rds("data/btw21-districts-centroid.rds")
```

# Votes · 🗳

## Results

party vote share federal level

+ candidate tier (first tier) — ("Erststimmen")
+ party tier (second tier) — ("Zweitstimmen")

```{r, message=FALSE}
btw_dt |>
  filter(level == "federal") |>
  mutate(party = if_else(party %in% c("CDU", "CSU"), "CDU/CSU", party)) |>
  group_by(votes_type, party) |>
  summarise(share = sum(share) |> round(1)) |>
  pivot_wider(names_from = votes_type, values_from = share)
```

## Party votes

```{r, message=FALSE}
#| fig-height: 7

pl_dt <-
  wk_shp |>
  select(wkr_nr, geometry) |>
  right_join(btw_tier2) |>
  mutate(party = if_else(party %in% c("CDU", "CSU"), "CDU/CSU", party)) |>
  group_by(party) |>
  mutate(share_diff = share - mean(share))
```

__party tier__ ("Zweitstimmen") vote share electoral districts

```{r}
#| fig-height: 7

pl1 <-
  ggplot() +
  geom_sf(data = pl_dt, aes(fill = share), lwd = 0.05) +
  scale_fill_viridis_c(option = "magma", direction = -1) +
  coord_sf(crs = crs_deu) +
  facet_wrap(vars(party))

pl1 + plot_annotation(caption = "Source: Bundeswahlleiter 2021")
```

__difference (+/-)__ of vote share party __district__ from __national average__ vote share party

```{r}
#| fig-height: 7

pl2 <-
  ggplot() +
  geom_sf(data = pl_dt, aes(fill = share_diff), lwd = 0.05) +
  scale_fill_gradient2(name = "+/-") +
  coord_sf(crs = crs_deu) +
  facet_wrap(vars(party))

pl2 + plot_annotation(caption = "Source: Bundeswahlleiter 2021")

pl_out <- pl1 / pl2 + plot_annotation(caption = "Source: Bundeswahlleiter 2021")

ggsave("z-btw21-vote-share.png", pl_out, width = 5, height = 7)
```

## Candidate votes

__candidate tier__ ("Erststimmen") vote share by party

+ A --- plurality vote __winning party__
+ B --- sum of candidate vote share __difference__ by __ideology__
  + right --- AfD, CDU, CSU, FDP
  + left --- DIE LINKE, GRÜNE, SPD
+ I -- vote share __plurality winner__
+ II -- vote share __difference top 2__ results


```{r, message=FALSE}
party_scale <- c("#A6CEE3", "#1F78B4", "#1b699e", "#E31A1C", "#33A02C", "#FB9A99")

ideo_share <-
  btw_tier1 |>
  group_by(wkr_nr, ideology) |>
  summarise(share = sum(share)) |>
  pivot_wider(names_from = ideology, values_from = share) |>
  mutate(rl_diff = right - left)

pl_dt <-
  wk_shp |>
  select(wkr_nr, geometry) |>
  inner_join(ideo_share) |>
  inner_join(mp |> select(wkr_nr, party))

pl_dt_ctr <- wk_ctr |> left_join(mp)

pl1 <-
  ggplot() +
  geom_sf(data = pl_dt, aes(fill = party), lwd = 0.05) +
  geom_sf(data = pl_dt_ctr, aes(shape = party), alpha = 0.4) +
  coord_sf(crs = crs_deu) +
  scale_fill_manual(values = party_scale)
# print(pl1)

pl2 <-
  ggplot() +
  geom_sf(data = pl_dt, aes(fill = rl_diff), lwd = 0.05) +
  coord_sf(crs = crs_deu) +
  scale_fill_gradient2(name = "right +/-")
# print(pl2)

pl1 + pl2 + plot_annotation(tag_levels = "A") +
  plot_annotation(caption = "Source: Bundeswahlleiter 2021")
```

```{r, message=FALSE}
diff_1_2 <-
  btw_tier1 |>
  arrange(wkr_nr, desc(share)) |>
  group_by(wkr_nr) |>
  filter(row_number() %in% c(1, 2)) |>
  summarise(
    top_2_diff = max(share) - min(share),
    share = max(share)
  )

pl_dt <-
  wk_shp |>
  select(wkr_nr, geometry) |>
  inner_join(diff_1_2)

pl1 <-
  ggplot() +
  geom_sf(data = pl_dt, aes(fill = share), lwd = 0.05) +
  coord_sf(crs = crs_deu) +
  scale_fill_viridis_c(option = "cividis", direction = -1)
# print(pl1)

pl2 <-
  ggplot() +
  geom_sf(data = pl_dt, aes(fill = top_2_diff), lwd = 0.05) +
  coord_sf(crs = crs_deu) +
  scale_fill_viridis_c(option = "magma", direction = -1)
# print(pl2)

pl1 + pl2 + plot_annotation(tag_levels = "I") +
  plot_annotation(caption = "Source: Bundeswahlleiter 2021")
```

# MPs (SMD) · 🗺

## Maps

plurality winner candidate (first) tier vote by party and sex

### Germany

```{r, message=FALSE}
pl_dt_shp <- wk_shp |> left_join(mp)
pl_dt_ctr <- wk_ctr |> left_join(mp)

pl1 <-
  ggplot() +
  geom_sf(data = pl_dt_shp, aes(fill = party), lwd = 0.1) +
  # scale_fill_brewer(type = "qual", palette = 4) +
  scale_fill_manual(values = party_scale) +
  geom_sf(data = pl_dt_ctr, aes(shape = sex), colour = "grey15", alpha = 0.7) +
  coord_sf(crs = crs_deu)

pl2 <-
  ggplot() +
  geom_sf(data = pl_dt_shp, colour = "lightgrey", fill = "grey95", lwd = 0.1) +
  geom_sf(data = pl_dt_ctr, aes(shape = sex, colour = party)) +
  scale_colour_manual(values = party_scale) +
  guides(colour = "none", fill = "none", shape = "none") +
  coord_sf(crs = crs_deu)

pl1 + pl2 + plot_annotation(caption = "Source: Bundeswahlleiter 2021")
```

### NRW and Berlin

```{r}
#| fig-height: 3

plot_state <- function(state_select, party_scale_select) {
  pl_dt_shp_re <- pl_dt_shp |> filter(state == state_select)
  pl_dt_ctr_re <- pl_dt_ctr |> filter(state == state_select)

  ggplot() +
    geom_sf(data = pl_dt_shp_re, aes(fill = party), lwd = 0.2) +
    scale_fill_manual(values = party_scale_select) +
    geom_sf(data = pl_dt_ctr_re, aes(shape = sex), colour = "grey30", size = 2) +
    guides(colour = "none", fill = "none", shape = "none")
}

pl1 <- plot_state("NW", party_scale[c(2, 4, 5)])
pl2 <- plot_state("BE", party_scale[c(2, 4, 5, 6)])
# plot_state("HH", party_scale[c(2, 5, 6)])
pl1 + pl2 + plot_layout(widths = c(55, 45)) +
  plot_annotation(caption = "Source: Bundeswahlleiter 2021")
```


## Descriptives

number (n) of districts won by party and sex (f -- female, m -- male)

### Party

```{r}
count_party <-
  mp |>
  count(party, sex) |>
  pivot_wider(names_from = sex, values_from = n) |>
  mutate(share_f = round(100 * f / (m + f)))

count_party |> arrange(share_f)
```

```{r}
mp |>
  count(party) |>
  mutate(share = round(100 * n / sum(n))) |>
  arrange(desc(share))
```


### States

```{r, message=FALSE, rows.print=20}
count_state <-
  mp |>
  left_join(wk_shp |> select(wkr_nr, state_name = land_name)) |>
  count(state_name, sex) |>
  pivot_wider(names_from = sex, values_from = n) |>
  mutate(share_f = round(100 * f / (m + f)))

count_state |> arrange(share_f)
```

# Models · 📈

+ area_km2 --- size of district as square length (root of area)
+ coord_y --- north/south position (latitude)
+ left_share --- share of left parties (SPD, Linke, Grüne)
+ region -- western or eastern (former GDR) state

```{r, message=FALSE}
library(broom)
library(ggeffects)

mo_dt <-
  pl_dt_shp |>
  left_join(ideo_share) |>
  mutate(
    female = if_else(sex == "f", 1, 0),
    left_share = left
  )
```

## Female first-tier winners

Logit model to predict female winners of candidate (first) tier vote

```{r, warning=FALSE}
mo_glm <- glm(female ~ area_km2 + coord_y + left_share, data = mo_dt, family = "binomial")

tidy(mo_glm) |> mutate(across(where(is.numeric), round, 2))

pl1 <-
  ggpredict(mo_glm, terms = "area_km2 [all]") |>
  plot(add.data = TRUE, show.title = FALSE)

pl2 <-
  ggpredict(mo_glm, terms = "coord_y [all]") |>
  plot(add.data = TRUE, show.title = FALSE, show.y.title = FALSE)

pl1 + pl2
```
## Left vote-share

Linear model of left parties (SPD, Linke, Grüne) vote share

```{r, warning=FALSE}
mo_lm <- lm_robust(left_share ~ area_km2 * region + coord_y * region, data = mo_dt)

tidy(mo_lm) |>
  mutate(across(where(is.numeric), round, 2)) |>
  select(-conf.low, -conf.high, -df, -outcome)


pl1 <-
  ggpredict(mo_lm, c("area_km2", "region")) |>
  plot(add.data = TRUE, show.title = FALSE, show.legend = FALSE)

pl2 <-
  ggpredict(mo_lm, terms = c("coord_y", "region")) |>
  plot(add.data = TRUE, show.title = FALSE, show.y.title = )

pl1 + pl2
```

## Party vote-share

Linear model of party vote share

```{r message=FALSE, warning=FALSE}
mo_dt2 <-
  btw_tier2 |>
  mutate(party = if_else(party %in% c("CDU", "CSU"), "CDU/CSU", party)) |>
  # mutate(party = fct_relevel(party, "CDU")) |>
  left_join(wk_shp)

mo_lm2 <- lm_robust(share ~ party * area_km2 * region, data = mo_dt2)

# tidy(mo_lm2) |> mutate(across(where(is.numeric), round, 2))

ggpredict(mo_lm2, terms = c("area_km2", "region", "party")) |>
  plot(add.data = TRUE, show.title = FALSE)
```
